How to Avoid OOTA Without Really Trying

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Remove Section 4 tagline
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Section 4.4 s/z/x/ in code example
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add "full C++" qualifier for last of Section 5.3
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\emph{Tagline: Compilers have considerable freedom, but are constrained
in optimizing atomics, especially volatile atomics.}

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which T0 loads the value of \co{x} and stores it in \co{y} while T1
does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code:
% Constructed by hand.
\begin{quote}
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~z = r2} \\
\end{tabular}
\end{quote}
The loads previously on line~1 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

@@

If all the stores in the OOTA cycle are omitted then all the reads must
also be omitted, meaning that the entire cycle has been optimized away.
We conjecture that in this situation there must be another abstract
execution which has the same observable effects as $A$ and is also
realized by $H$, but in which the OOTA cycle does not occur.
Thus there would be no way to tell, merely by observing the effects of
$H$, whether or not there was an OOTA cycle.
We therefore declare OOTA cycles in which all accesses are omitted to
be \emph{trivial}.

>>>>

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which T0 loads the value of \co{x} and stores it in \co{y} while T1
does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code:
% Constructed by hand.
\begin{quote}
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}
\end{quote}
The loads previously on line~1 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

@@

If all the stores in the OOTA cycle are omitted then all the reads must
also be omitted, meaning that the entire cycle has been optimized away.
We conjecture that in this situation for full C++ there must be another
abstract execution which has the same observable effects as $A$ and is
also realized by $H$, but in which the OOTA cycle does not occur.
Thus there would be no way to tell, merely by observing the effects of
$H$, whether or not there was an OOTA cycle.
We therefore declare OOTA cycles in which all accesses are omitted to
be \emph{trivial}.
